The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2021-07-14 Rmd/trajectories_manuscript.tex
First corrections to Núria's comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo García-Castro1, Daniela Avila-Varela1, \& Núria Sebastian-Galles1},
  pdflang={en-EN},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{The role of cross-linguistic lexical similarity on bilingual word acquisition}
\author{Gonzalo García-Castro\textsuperscript{1}, Daniela Avila-Varela\textsuperscript{1}, \& Núria Sebastian-Galles\textsuperscript{1}}
\date{}

@@

Learning words involves two key steps: (1) the encoding of a word form (e.g, /dɒɡ/)
and (2) its association with a referent (e.g., a DOG). There is evidence that word learning occurs at early ages: six-month-old infants show a preference toward named pictures, relative to unnamed pictures displayed side-by-side ({\textbf{???}}). At 7.5 months, the phonological representations of familiar words seem to be quite detailed, as infants show evidence of recognition when the whole word is uttered ({\textbf{???}}), as opposed to when only part of the word is uttered. At 12 months, infants show sensitivity to both vowel and consonant mispronunciations ({\textbf{???}}), looking shorter to named pictures when their label is mispronounced. Remarkable as it is, the phonological specificity of infants' lexical entries is only the foundation of following developmental milestones. One of them, central to this study, is the emergence of excitatory and inhibitory links between word representations, which represents one of the essential characteristics of the adult lexicon ({\textbf{???}}).

Words are not learnt in isolation, but in contexts rich in word tokens and referents. The structure of toddlers' lexicon reflects this high connectivity ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}). At 18 months, infants' recognition of spoken words is sensitive to phonological priming effects, suggesting that their lexical entries are phonological linked, and co-activated during speech processing ({\textbf{???}}; {\textbf{???}}). The emergence of semantic links in the developing lexicon occurs somewhat later, at 24 months of age, when infants show sensitivity to semantic priming ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and show evidence of inhibitory links between lexical-semantic entries ({\textbf{???}}). By 24 months, toddlers' recognition of familiar words also seems to follow a hierarchical fashion, as revealed by their sensitivity to the phonology and semantic similarity of distractor pictures presented along the named picture: the interference of phonological distractors is stronger at earlier stages of spoken word recognition, that of semantic distractors is stronger at later stages ({\textbf{???}}). Vocabulary size predicts the emergence and the strength of these effects: 18 months-old toddlers are sensitive to semantic priming and interference too, provided they know a similar amount of words ({\textbf{???}}; {\textbf{???}}), and the strength of the phonological and semantic distractors during spoken word recognition grows stronger along the vocabulary size of the toddlers. This suggests that the emergence of rich lexical networks -- both at the phonological and the semantic levels -- grows as a function of the number of words acquired by the toddler.

The structure of the lexicon also impacts the order of acquisition of new words. Words that share a high degree of phonological and/or semantic overlap with words already acquired are more likely to be acquired next ({\textbf{???}}). ({\textbf{???}}) and ({\textbf{???}}) showed that the connectivity of a given semantic category (i.e., animals) in a child's lexicon predicts a better performance in a disambiguation task, where participants are presented with a novel label in the presence of a familiar and a novel object: infants show stronger looking preference for the novel object if it belonged to a category for which many words had already been acquired. This points to the structure of the lexicon facilitating the strategies children engage during word learning.

The case of bilingual children (here defined as those learning two languages from birth) presents an opportunity to study how lexical acquisition takes placed under a more complex environment, as they face the challenge of learning two distinct sets of words -- one for each language -- that partially overlap in sound and meaning.

EARLY STAGES IN BILINGUAL LANGUAGE ACQUISITION

When compared to their monolingual peers, they know fewer words only when just on language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When both languages are taken into account, bilingual children seem to know, at least, as many words as monolinguals do (Ben-Zeev, 1977; Bialystok, Luk, Peets, \& Yang, 2010; Blom et al., 2019; Core, Hoff, Rumiche, \& Señor, 2013; Doyle \& others, 1977; Fernandez, Pearson, Umbel, Oiler, \& Molinet-Molina, 1992; Hoff et al., 2012; Hoff, Rumiche, Burridge, Ribot, \& Welsh, 2014; Pearson \& Fernández, 1994; Rosenblum \& Pinker, 1983; but see Pearson, Fernández, \& Oller, 1993; Houwer, Bornstein, \& Putnick, 2014).

Floccia et al. (2018), gathered parental vocabulary estimates ({\textbf{???}}) from a large sample of 24-month-old toddlers learning English and an additional language (from a pool of 13 linguistically diverse languages). They computed the average phonological similarity between translation equivalents (TEs from now on; e.g., \emph{table} in English and \emph{tafel} in Dutch) for each pair of languages, and found a positive association between this measure of phonological similarity and participants' productive vocabulary sizes in their additional language. Using a similar measure of lexical similarity, Blom et al. (2019) extended these results to bilingual children aged three to 10, and reported a positive association between the lexical distance and comprehensive vocabulary size: bilinguals learning lexically close languages showed similar vocabulary sizes to those of monolinguals, but those learning the more distant languages showed lower vocabulary sizes.

These results suggest that linguistic distance between languages plays an important role during bilingual lexical acquisition, and that this effect operates within TEs. What mechanisms underlie this effect, and why they seem to operate differently across comprehension and production, or across the dominant and the non-dominant language, remain open issues. Floccia et al. (2018) pointed to parallel activation as a candidate mechanism behind this facilitation effect. The parallel activation principle suggests that bilinguals activate lexical representations in a language non-selective way: during the comprehension or production of a word in one language (e.g., \emph{cat}, in English) its corresponding translation in the other language is activated too {[}e.g., \emph{gato}, in Spanish. There is a vast body of evidence supporting language non-selective lexical access in both adults (e.g., Dell \& O'Seaghdha, 1991; {\textbf{???}}; Bobb, Von Holzen, Mayor, Mani, \& Carreiras, 2020; Dijkstra, 2005; Hoshino \& Kroll, 2008; Kroll, Gullifer, \& Rossi, 2013; Singh, 2014; Spivey \& Marian, 1999; Thierry \& Wu, 2007; Von Holzen, Fennell, \& Mani, 2019; Yudes, Macizo, \& Bajo, 2010; see Costa, Santesteban, \& Caño, 2005 for review), and children (e.g.~Von Holzen \& Mani, 2012; Bosma \& Nota, 2020; Jardak \& Byers-Heinlein, 2019; Poulin-Dubois, Bialystok, Blaye, Polonia, \& Yott, 2013). A critical implication of parallel activation is that the word form (i.e., phonology, orthography, etc.) of both translation equivalents are available to each other, impacting the comprehension or production dynamic of any of them.

Floccia et al. (2018) argue that the degree of phonological similarity between TEs (\emph{cognateness} from now on) of both languages should lead to increased parallel activation of both languages during language exposure, facilitating the acquisition of words in both languages, and leading participants leading two lexically close languages to show larger vocabulary sizes than those learning two lexically distant languages. This account, however, neglects the fact that for parallel activation to take place and play a facilitation role, lexical representations of both pairs of the TE must have been already established in the lexicon, and their corresponding phonological forms must have been encoded. This is not the case during lexical acquisition, where at least one of the members of the pair has not being acquired yet. A more lenient definition of parallel activation could adjust to the context of lexical acquisition more easily: when only one of the members of the TE has been acquired, its lexical representation is activated when, in the presence of its referent, a phonologically similar label is uttered. Consider the case of a Catalan-Spanish bilingual that has already learned the word \emph{gat} (Catalan for \emph{cat}), but not \emph{gato} (in Spanish). It is possible that, when presented with \emph{gato} in the presence of a cat, she maps this novel label to the familiar word \emph{gat}, and creates a lexical representation for \emph{gato} as a synonym for \emph{gat}. The case of a non-cognate is differing. Consider now the case of a toddler that has already learnt \emph{gos} (Catalan for \emph{dog}), but not \emph{perro} (Spanish). In this case, it would not be possible to map both labels via phonology, give their lack of phonological similarity.

Under this account, cognates (i.e., phonologically similar TEs) would be acquired earlier than non-cognates, but this effect would only play a role once one of the members of the TE has been acquired. The evidence supporting an earlier age of acquisition of cognates relative to non-cognates if sparse. Schelletter (2002)`s reported a longitudinal case-study of a German/English child who produced cognate TEs earlier than non-cognate TE. Bosch and Ramon-Casas (2014), showed converging results from a sample of 48 monolingual and bilingual infants aged 24 months. The low number of words included in the analyses of both studies, and the lack of adjustment for lexical frequency, limits severely the strength of any conclusions concerning the difference in age of acquisition of cognates and non-cognates. On the other hand Floccia et al. (2018) worked with aggregated estimates of participants' vocabulary size, losing item-level information that could provide information about change in age of acquisition of TEs associated with their degree of phonological similarity. Therefore, their results can only be considered as compatible with cognateness playing a role on age of acquisition. In this study, we aimed at overcoming these pitfalls by using unaggregated data from individual responses, testing the role of cognateness explicitly and adjusting for lexical frequency.

It also follows from the fact that cognateness operates \emph{within} TEs that the effect of cognateness can only play a role once one of the members of the TE has been acquired, and its phonological form is available during the acquisition of its translation. Words that belong to the language of most exposure (dominant language) are likely to be acquired earlier than words in the non-dominant language (Floccia et al., 2018). ({\textbf{???}}) tested Dutch-Frisian bilingual children aged 2.5 to four years in a comprehension task (PPVT-NL) that involved Frisian words with different degrees of phonological similarity with their Dutch TE. Infants with lower exposure to Frisian showed a better performance for words with high similarity, while no such benefit was found in children who were exposed mostly to Frisian. This suggests that infants that were mostly exposed to Dutch used the phonology of dutch words when processing Frisian words, as revealed by a better performance for cognates than for non-cognates.
A study by ({\textbf{???}}) showed that the degree of phonological similarity between
the performance of was better for cognates than for non-cognates, only for those children

Finally, previous studies have reported that the properties that describe the form of a word play a stronger role in production than comprehension than in production. This is the case of the number of phonemes (Braginsky, Yurovsky, Marchman, \& Frank, 2019), phonological neighbourhood density (Jones \& Brandt, 2019). These results converge with Floccia et al. (2018)'s finding that the effect of the average cognateness between languages was larger in production than in comprehension\footnote{Although Bosch and Ramon-Casas (2014) found that participants in their sample produced cognates earlier than non-cognates, no analyses were conducted on comprehensive data, and therefore they cannot be taken as supporting evidence for a larger effect of cognateness on production than in comprehension}. Accordingly, we expected the role of cognateness to be more central in production than in comprehension.

In summary, we investigated the effect of cognateness on the probability of acquisition of TEs, and predicted that (1) cognate TEs are acquired earlier than non-cognate TEs, that this effect will be larger in the dominant than in the non-dominant language, and larger in production than in comprehension.

@@

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

We collected data from 349 bilinguals (167 female), from the Metropolitan Area of Barcelona, between 28th October, 2019 and 09th January, 2021. All families gave informed consent before participating. This study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX. We assessed toddlers' language profile asking parents for an estimated proportion of exposure to each language. We excluded participants with \textgreater10\% exposure to a third language. Fig. \ref{tab:lps} illustrates the distribution of participants across language profiles and ages.

\hypertarget{questionnaire}{%
\subsection{Questionnaire}\label{questionnaire}}

We implemented an on-line questionnaire using \texttt{formr} (Arslan, Walther, \& Tata, 2020), divided in three forms: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (Hamilton, Plunkett, \& Schafer, 2000) and consisted in two lists of words, one in Catalan and one in Spanish). Items in one language were translation equivalents of the items in the other (e.g., whenever \emph{gos} {[}dog{]} was included in the Catalan inventory, the word \emph{perro} was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both (e.g., Catalan \emph{acabar} {[}\emph{to finish}{]} and Spanish \emph{acabar} and \emph{terminar}), or merged them into the same one (e.g., Spanish \emph{mono} {[}\emph{monkey}{]} and Catalan \emph{mono/mico}). We included items from a diverse sample of 26 semantic/functional categories (see Appendix 1). We discarded the following categories: adverbs, auxiliary words, connectives, interjections and games and routines. The Catalan inventory contained 778 items (196 cognates, 582 non-cognates) and the Spanish inventory contained 781 (197 cognates, 584 non-cognates).

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand and say it, or did not understand or say it (marked by default). Participants filled a long or a short version of the questionnaire. Participants presented with the long version filled a list of 800 translation equivalents (800 items in Catalan and 800 items in Spanish), while participants presented with a short were randomly allocated into one of four list of items. Each list contained a different set of \textasciitilde400 translation equivalents (\textasciitilde400 in Catalan, \textasciitilde400 in Spanish). Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists - were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., \emph{barrita de cereales} {[}cereal bar{]}) and items that included more than one word-form (e.g., \emph{mono / mico}). Table \ref{tab:itemsinfo} shows the classification of items in cognates and non-cognates and their frequency scores across the four lists of the inventories.

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

We gathered 181587 responses. Translation equivalents received an average of 386.36 (\emph{SD} = 228.76, , Max = 698), both languages summed together. We modelled the probability of responses (\(y_i\)) to belong to any of the three possible categories (\(k\)): 1 = \emph{No} or say, 2 = \emph{Understands}, and 3 = \emph{Understands and Says}) using an ordinal regression model with a cumulative logit link function (Agresti, 2010; Bürkner \& Vuorre, 2019). This model estimates the probability of response \(y_i\) to belong to category \(k\) or lower. In our case, this will result in the estimation of the probability of \(y_i \leq 1\), \(y_i \leq 2\), and \(y_i \leq 3\), where

\[
p_k = Pr(y_i = k) = Pr(y_i \leq k) - Pr(y_i \leq k-1).
\]

We include several predictors of interest to adjust this probability to participants' and items' (lexical frequency, dominance, cognateness) characteristics. These predictors are:

\begin{itemize}
\tightlist
\item
  Age of participant in months (\texttt{age}, \(Age\)) calculated as the difference in days between participants' birth date and questionnaire completion divided by 30 and centred around the mean
\item
  Degree of bilingualism of participant (\texttt{bilingualism}, \(Bilingualism\)) computed as the percentage of exposure to a second language (Spanish for participants exposed to \textgreater50\% to Catalan an \emph{vice versa}) rounded to units and centered around the mean
\item
  Lexical frequency of item (\texttt{frequency}, \(Frequency\)) retrieved from SUBTLEX-CAT (Boada, Guasch, Haro, Demestre, \& Ferré, 2019) for Catalan words and from SUBTLEX-ESP (Cuetos, Glez-Nosti, Barbón, \& Brysbaert, 2011) for Spanish words, expressed as Zipf scores (Heuven, Mandera, Keuleers, \& Brysbaert, 2014), and centred around the mean
\item
  Language the item belongs to (\texttt{dominance}, \(Dominance\)) labelled as \texttt{L1} if the item belongs to the dominant language (e.g., a Catalan item is labelled as \texttt{L1} for Catalan-dominant participants and as \texttt{L2} for Spanish-dominant participants), coded as \texttt{L2} = -0.5 and \texttt{L1} = 0.5
\item
  Cognateness of the item (\texttt{cognate}, \(Cognate\)), coded as \texttt{Non-cognate} = -0.5 and \texttt{Cognate} = 0.5
\end{itemize}

We adopted a Bayesian approach toward statistical inference, which allows to (1) incorporate previous domain knowledge into the inference process implementing the Bayes theorem, and (2) to quantify the uncertainty associated to the estimated parameters in our model. Our extended model can be formalised as follows:

\[
\begin{aligned}
y_i \sim Ordered(p) \\
log\frac{Pr(y_i \leq k)}{1-Pr(y_i \leq k)} &= \alpha_k - \dots \\
& \beta_1 \times Age_i + \dots \\
& \beta_2 \times Frequency_i + \dots \\
& \beta_3 \times Dominance_i + \dots \\
& \beta_4 \times Bilingualism_i + \dots \\
& \beta_5 \times Cognate_i + \dots \\
& \beta_6 \times (Dominance_i \times Bilingualism_i) + \dots \\
& \beta_7 \times (Dominance_i \times Cognate_i) + \dots \\
& \beta_8 \times (Bilingualism_i \times Cognate_i) + \dots \\
& \beta_9 \times (Dominance_i \times Bilingualism_i \times Cognate_i), \\
\alpha \sim Normal(0, 10) \\
\beta_{0-9} \sim Normal(0, 10)
\end{aligned}
\]

Where:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo Garcia-Castro1, Daniela Avila-Varela1, \& Nuria Sebastian-Galles1},
  pdflang={en-EN},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{The role of cross-linguistic lexical similarity on bilingual word acquisition}
\author{Gonzalo Garcia-Castro\textsuperscript{1}, Daniela Avila-Varela\textsuperscript{1}, \& Nuria Sebastian-Galles\textsuperscript{1}}
\date{}

@@

\hypertarget{introduction-1}{%
\section{Introduction}\label{introduction-1}}

Before the end of their first year of life, infants start directing their gaze to some object when hearing their labels, according to both experimental data (Bergelson \& Swingley, 2012; Jusczyk \& Aslin, 1995; Tincoff \& Jusczyk, 1999) and parental reports (e.g., Fenson et al., 1994). During the last half of their second year, they acquire new words at an increasingly fast rate (Bergelson, 2020; Bloom, 2002; Fenson et al., 1994). Despite the considerable variation in children's trajectories of vocabulary growth, such trajectories seem quite consistent across languages (Braginsky, Yurovsky, Marchman, \& Frank, 2019). Nonetheless, most of the literature on early word acquisition has been conducted on monolingual children, and does not address the problem of how bilinguals--who represent a substantial proportion of the population in most societies--acquire words at early ages.

There is evidence that bilinguals know less words in each of their languages than monolinguals, but also that both groups know a similar amount of words when the two languages are aggregated. For example, Hoff et al. (2012) found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals (who only learnt English). In contrast, both groups knew a similar total amount of words, when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated (Fabian, 2016; Gonzalez-Barrero, Schott, \& Byers-Heinlein, 2020; Oller \& Eilers, 2002; J. L. Patterson, 2004; J. Patterson, Pearson, \& Goldstein, 2004; Pearson \& Fernández, 1994; Pearson, Fernández, \& Oller, 1993; Petitto et al., 2001; Smithson, Paradis, \& Nicoladis, 2014). While these studies have mostly relied on samples of bilingual children learning two relatively distant languages (as it is the case of English and Spanish) it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms--one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in varying degrees. Particularly, when both languages are linguistically close, like Spanish and Catalan (both Roman languages), they are more likely to share a larger amount of cognates (i.e., form-similar translation equivalents) than two linguistically distant languages, like Spanish and English--one Roman, the other Germanic. For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear \emph{puerta} and \emph{porta} (cognates), whereas a Spanish-English bilingual might hear \emph{puerta} and \emph{door} (non-cognates). It could be the case that mapping two phonologically similar labels (cognates, like \emph{puerta}-\emph{porta}) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like \emph{puerta} and \emph{door}). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

A recent study by Floccia et al. (2018) provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers in the United Kingdom who were learning English and an additional language. The additional language was one a pool of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English. Phonological similarity was measuring by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the edit distance between two strings by counting the smallest number of insertions, deletions and substitutions one of the strings has to go through to become identical to the other (Levenshtein \& others, 1966). Floccia et al.~divided the resulting score by the length of the longest string to bound the similarity scores between 0 and 1, and then entered this variable as a predictor as they modelled participants' vocabulary sizes. Among other findings, the authors reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14\% phonological similarity), were able to produce more Dutch words than English-Mandarin bilinguals (1.97\% phonological similarity) were able to produce in Mandarin.

Floccia et al.~pointed to \emph{parallel activation} as the main mechanism underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension. This phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production (Costa, Caramazza, \& Sebastian-Galles, 2000; Hoshino \& Kroll, 2008) or comprehension (Thierry \& Wu, 2007). One of the clearest examples of parallel activation was provided by Costa, Caramazza, and Sebastian-Galles (2000). In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre}, translations of \emph{tree}), whereas in the other half of the trials labels were non-cognates (\emph{mesa}-\emph{taula}, translations of \emph{table}). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals did not show this effect. These results suggest that, for the bilinguals, Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several recent studies have also provided similar evidence in comprehension in children (Poulin-Dubois, Bialystok, Blaye, Polonia, \& Yott, 2013; e.g., Von Holzen \& Mani, 2012). Parallel activation is therefore a plausible mechanism to account for Floccia et al.'s results: cognates increase the amount of activation in both languages, facilitating word acquisition. Therefore, children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. If cognates are the reason for this facilitation effect, it could be predicted that cognates are acquired earlier than non-cognates. Likely as it might seem, an earlier age of acquisition for cognates is not the only scenario where Floccia et al.'s results are account for.
For example, languages that share more cognates might also share a larger overlap in their phonemic inventories, which could facilitate the acquisition of all words, regardless of their cognate status.

However, the evidence on an earlier age of acquisition for cognates compared to non-cognates is scarce. To the best of our knowledge only two studies have provided direct data on this issue. First, Bosch and Ramon-Casas (2014) used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. Second, Schelletter (2002) reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these studies make it challenging to draw strong conclusions about the effect of cognateness in vocabulary growth. Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp.~70):

\begin{quote}
``This finding also provides support to the proposal that the cognate advantage is due to cognates being acquired before non-cognates in early childhood (Costa et al., 2016), leading to an ease of processing later in life.''
\end{quote}

We identify two main reasons why we cannot conclude an earlier age of acquisition for cognates than for non cognate from Floccia et al.'s results. First, the response variable used was the proportion of words each participant understood and/or produced, from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs.~non-cognates was lost. Second, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at this age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item. In summary, the evidence supporting an earlier age of acquisition of cognates vs.~non-cognates presents some limitations that prevents drawing sound conclusions about this effect.

Another worth-mentioning finding in Floccia et al.~is that the increase in vocabulary size associated with linguistic similarity was larger in the additional language vocabulary than in English vocabulary. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Given that the relative amount of exposure to a given language is a strong predictor of children's vocabulary size in that language (e.g., Cattani et al., 2014; Thordardottir, Rothenberg, Rivard, \& Naves, 2006), participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language. If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

The aim of this study is to investigate the role of cognateness in bilingual word acquisition. We hypothesised that cognate words would be acquired earlier than non-cognate word, and that this difference would be larger in participants with lower exposure to the language such word belongs to. Using an designed online vocabulary checklist--carefully designed specifically for this study--we collected data from a sample of bilinguals aged 10 to 36 months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then modelled the probability of participants being reported by their parents to understand and/or say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for participants' age and the item's lexical frequency.

\hypertarget{references}{%
\subsection{References}\label{references}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{references-1}{%
\section{References}\label{references-1}}
